Massively Parallel Pseudo-Spectral DNS and~LES for Particle-Laden Turbulent Flows under Two-Way Momentum Coupling

### 2023-02-04 msc-thesis.tex
removed unnecessary package references
some typographic adjustments for print

\usepackage[table, svgnames, dvipsnames]{xcolor}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage[framemethod=tikz]{mdframed}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{epsfig}
\usepackage{pstricks}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage[backend=biber,style=authoryear-comp,sorting=nyt,maxcitenames=2,uniquename=false,uniquelist=false,maxbibnames=5,maxsortnames=1]{biblatex}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage[hidelinks,allcolors=black]{hyperref}
\usepackage{upgreek} 
\usepackage{placeins}
\renewcommand{\arraystretch}{1.1}

@@

\keywords{obliczeniowa mechanika płynów, przepływ turbulentny, przepływ wielofazowy, krople, mikrofizyka chmur, obliczenia dużej mocy, równania Naviera-Stokesa, homogeniczna izotropowa turbulencja, statystyki zderzeniowe kropel, metoda pseudo-spektralna, bezpośrednia symulacja numeryczna (DNS), metoda dużych wirów (LES), dwustronne sprzeżenie pędu, efekty grawitacyjne}

@@

\streszczenie{
Przepływy turbulentne z cząstkami fazy rozproszonej zachodzą nieustannie w środowisku naturalnym oraz znajdują liczne zastosowania w procesach przemysłowych.
Ważnym tego przykładem są chmury atmosferyczne, które składają się z ogromnej ilości małych kropel wody i kryształków lodu.
Wyjaśnienie wzajemnego oddziaływania powietrza (płynu) i aerozolu (cząstek) jest istotne w kontekście poznania mechanizmów rządzących powstawaniem i przebiegiem zjawisk atmosferycznych.
Wiedza w tym zakresie jest użyteczna w obszarze meteorologii i służy do opracowywania coraz to bardziej realistycznych modeli powstawania opadu z kropel chmurowych.
To z kolei prowadzi do podniesienia dokładności i sprawdzalności numerycznych prognoz pogody.
Jednym ze sposobów badania przepływów dyspersyjnych są symulacje numeryczne. 
\newline \indent
Powszechnie stosowanym do tego celu narzędziem, zapewniającym wysoką dokładność, są tzw. bezpośrednie symulacje numeryczne (direct numerical simulation, DNS). 
Za pomocą symulacji DNS można pozyskiwać statystyki zderzeniowe cząstek, które to są użyteczne jako parametry w modelach obejmujących większe skale przestrzenne.
Symulacje DNS są jednak kosztowne obliczeniowo, co ogranicza ich stosowalność do siatek o średnich rozmiarach (${\sim 2048^3}$ węzłów -- przy współcześnie dostępnych zasobach obliczeniowych).
DNS nie pozwalają modelować turbulencji charakteryzowanej wyższymi liczbami Reynoldsa ($R_{\lambda} \sim 10^4$), czyli takiej która obserwowana jest w powietrzu atmosferycznym. \newline \indent
Metoda dużych wirów (large-eddy simulation, LES) stanowi alternatywę dla DNS.
W tej metodzie drobnoskalowa turbulencja nie jest bezpośrednio rozwiązywana, ale jest parametryzowana przy użyciu modelu podskalowego.
Umożliwia to modelowanie przepływów turbulentnych o większej liczbie Reynoldsa przy użyciu siatek o mniejszej liczbie węzłów.
Ograniczenie złożoności obliczeniowej odbywa się jednak kosztem mniejszej precyzji i fizycznej dokładności.
Niniejsza praca stanowi porównanie obu metod (DNS i LES), zarówno w kwestii fizycznej dokładności, jak i wydajności, skupiając się w szczególności na symulacjach z dwustronnym sprzężeniem pędu pomiędzy płynem i cząstkami. 
Zostało wykazane, że LES stanowi obiecującą odpowiedź na ograniczenia DNSu, pomimo pewnych niedokładności wynikających z filtrowania turbulencji w najmniejszych skalach.
Różnice te, w większości przypadków, stają się mniejsze, gdy uwzględni się dwustronne sprzężenie pędu, w szczególności dla cząstek poddanych działaniu grawitacji.
\newline \indent
Przeprowadzone zostały również analizy kosztów obliczeniowych obu metod.
Choć zastosowanie LES zapewnia ogromną oszczędność w obliczeniach związanych z turbulentnym przepływem płynu, to jednak koszt ten znacząco rośnie wraz z liczbą cząstek.
Niezbędna ilość symulowanych cząstek może zaś być duża, gdyż efekty dwustronnego sprzężenia pędu objawiają się wyraźniej w układach z względnie wysokim stosunkiem masy cząstek i płynu (pomiędzy $0.1$~i~$1$).
Obliczenia odnoszące się do dynamiki cząstek stanowią więc bardziej istotne ograniczenie dla LESu niż dla DNSu, gdzie złożoność obliczeń związanych z~płynem jest czynnikiem dominującym.
Ponadto zidentyfikowano i przeanalizowano szereg parametrów, które pośrednio wpływają na czas trwania symulacji, takie jak np. promienie kropel oraz~działanie grawitacji.
Na koniec przedstawiono także wpływ preferencyjnego koncentrowania się cząstek na wariancję ich rozkładu pomiędzy równolegle wykonywanymi procesami.
Ten czynnik ma również istotne znaczenie w kontekście całkowitej wydajności symulacji.
}

@@

Many processes that involve fluid flows, whether laminar or turbulent, are composed of more than one phase.
Here, by phase, we understand regions of the system with the same properties that can be clearly separated by some interface from others.
We refer to such flows as multi-phase, or two-phase, in a common case where the number of considered phases is limited to two.
These phases often are of different states of matter, e.g. gas--liquid for air bubbles flowing in the water pipe, but systems with two immiscible liquids, such as water and crude oil in pipelines, are also studied \parencite{Brauner2003}.
In many cases, one of the phases may be treated as continuous (referred to as the continuous phase, or the carrier fluid), while the other remains in separated chunks.
We refer to such chunks as the disperse phase, and examples may include air bubbles, water droplets, solid particulate matter suspended in a fluid, etc.
Often enough, elements of disperse phase may be approximated as point particles, i.e. material points that are conceptually representing small spheres of given radii, that are considered insoluble and immiscible (that is, they cannot exchange mass with the carrier fluid, suddenly appear, or be destroyed).
Such approximation is especially accurate when disperse phase consists of small droplets or bubbles of one fluid surrounded by another, as the surface tension naturally preserves spherical shape of disperse elements.
Cloud droplets, in particular, remain spherical since the capillary pressure is several orders of magnitude larger than the local fluid shear stress induced by the disturbance flow associated with their motion.
Furthermore, \textcite[Figure 1, therein]{Balachandar2010} point out that for such approximation to be valid, particle response times have to be greater than Kolmogorov time scale of background turbulent flow. 
We shall refer to two-phase flows with disperse phase approximated by point particles as particle-laden.

@@

The modelling of atmospheric clouds is a broad domain that has to deal with wide range of length scales, where different structures and physical processes are important and have to be directly represented, while others need to be parameterised due to constraints in computing power.
Such tradeoff between computational complexity and physical fidelity of simulations gives rise to the plethora of methods used depending on the respective scale.
Simplified picture of that situation is presented in Figure \ref{fig:atmo-scales}.
At the largest scales we have global climate models (GCM) and global weather models (GWM, e.g. Integrated Forecasting System, Global Forecasting Systems, ICON), where computation domain  often spans the entire Earth ($\sim 10^{7}$~m) and each cell has (horizontal) dimensions ranging from $10$ to $100$ kilometres.
We also consider mesoscale models that are mainly used for weather prediction in smaller domains that usually cover regions, countries, or continents ($\sim 10^{5} \text{--} 10^{6}$ m) and grid spacing is in the order of kilometres.
For instance, maximal resolution of mesoscale model COSMO used for weather forecasting in Poland is $2.8$ km \parencite{Doms2021}.
In this context it is worth adding that a small cloud with a volume of $1 \; \text{km}^3$ may contain $10^{17}$ or more droplets.
Both types of models, due to relatively large grid cell sizes, cannot fully resolve even fairly large cloud structures and depend on a lot of parameterisations.
In the middle ground, we have cloud resolving models (CRM; domain sizes---kilometres, cell sizes---metres) that cannot fully resolve all turbulent scales of air or exactly model behaviour of every droplet, thus necessary parameterisations are employed, such as large-eddy simulations (LES) and super-droplet method \parencite{Arabas2013}.
CRM is commonly employed for research studies focused on the effects of microphysical processes affecting entire clouds, their formation and evolution, precipitation, etc.
For that reason, CRMs are used to refine cloud parameterisations used in GCM and mesoscale models.

\begin{figure}[h]
\centering
\includegraphics[width=13cm]{figures/0-01_atmo-scales.png}
\caption{
Hierarchy of atmospheric models and the scales of atmospheric motion reproduced from \textcite{Morrison2020} with author's permission.
Coloured rectangles represent types of simulations commonly used when working with respective length scales, and comments emphasise transition from directly resolving most elements of the model (left) to including more and more parameterisations with the increase of scale (right).
Magenta rectangle positioned on the axis was added to emphasise range of scales that corresponds to simulations discussed in this thesis.
}
\label{fig:atmo-scales}
\end{figure}

@@

Recent studies, however, have shown that direct numerical simulations are close to their limits due to prohibitively large computational costs.
\textcite{Rosa2013,Rosa2016} used grids of up to $1024^3$ nodes, which seems to be the reasonable limit for most studies, although there were studies using even larger grids---e.g. $2048^3$ nodes in \textcite{Ireland2016} and \textcite{Matsuda2021}; for flows without particles grids of size $8192^3$ were used \parencite{Buaria2019} and more recently even up to $12\,288^{3}$ \parencite{Buaria2020}.
It is important to note that grid size is crucial for turbulent flow simulations since it affects the range of spatial scales that can be resolved---from the smallest eddies with sizes close to the spatial grid spacing, to the largest ones of sizes typically about $1/3$ of the computational domain (this limitation is due to a numerical requirement that the correlation of fluid velocity at distances comparable to the domain size must vanish).
Such range between largest and smallest resolvable scales can be expressed using Reynolds numbe, which is commonly interpreted as a measure of magnitude of turbuence.
Examples of values of the Taylor microscale (that is, considering only eddies in inertial subrange) Reynolds number, $R_{\lambda}$, attained in previous high-resolution studies are $R_{\lambda} = 499$ for $1024^3$ grid \parencite{Rosa2013,Rosa2016}; $R_{\lambda} = 597$ \parencite{Ireland2016}, and $R_{\lambda} = 531$ \parencite{Matsuda2021} for $2048^3$ grids; $R_{\lambda} = 650$ for $8192^3$ \parencite{Buaria2019}; and $R_{\lambda} = 1300$ for $12\,288^3$ grid.
It is desirable, however, to attain higher values of $R_{\lambda}$ to simulate conditions that are present in typical atmospheric clouds, where $R_{\lambda}$ is known to attain values between $10^{3}$ and $10^{4}$.
What is more, the increase of the Taylor microscale Reynolds number with the number of grid nodes in one dimension, $N$, is sublinear (comparison of simulation data from various studies shows that $R_{\lambda} \propto N^{2/3}$, see \cite{Wang2009}, Figure 1 therein, which agrees with theoretical estimates for pseudo-spectral method in \cite{Pope2000}, Equation 9.8 therein).
Thus, when we double $N$ we only increase $R_{\lambda}$ by at most $50\% \text{--} 60\%$, while total number of grid nodes, $N^3$, increases by the factor of $8$.
The strategy of increasing grid sizes of DNS to model somewhat wider range of turbulent scales currently faces not only resource limitations, but is also hindered by the diminishing returns in the long run.
This situation calls for an alternative solution.

@@

Firstly, the accuracy, or physical fidelity, of LES is analysed.
Establishing exact measures of accuracy for collisional statistics is a challenging task because of experimental constraints, as simultaneous measurement of positions and velocities of large amounts of small droplets faces technical limitations, although considerable advances have been made recently in that field \parencite{Yavuz2018}.
Moreover, homogeneous and isotropic model of turbulence is an idealisation that could not be exactly replicated in actual physical conditions.
For these reasons it is difficult to obtain experimental results for validation, and due to the complexity of the problem at hand, no analytical solutions are available as well.
On the other hand, difficulties in obtaining experimental data makes estimations of these statistics \emph{in silico} even more valuable.
Thus, as in previous studies, the accuracy comparison will mostly focus on qualitative and quantitative differences between results obtained by DNS and LES.
The results of DNS are used as a baseline, under the assumption that this method is thoroughly studied and frequently used, and thus provides accurate estimates of parameters in question.
Hence, results obtained using LES are compared against them, in order to consider feasibility of LES as a workable replacement for DNS.
The further analysis will focus mainly on simulations under two-way momentum coupling.

@@

The numerical integration of the Navier-Stokes equations is performed using the -spectral method which is well-established for such simulations.
It was first introduced by \textcite{Orszag1972} and later developed with parallel execution in mind and for application to particle-laden flows (e.g. see \textcite{Wang1993,Ayala2014,Parishani2015}).
It involves transformations of the velocity field between its representations in physical space, $\textbf{U}(\textbf{x}, t)$, and spectral (Fourier) space, $\hat{\textbf{U}}(\textbf{k}, t)$, in order to perform some operations in the space that is most favourable.
Detailed description of that method may be found in Appendix \ref{app:psm}.
Note, that performing Fourier transforms (implemented as FFT) is the most computationally demanding part of the method.
In the version of pseudo-spectral method used here, every iteration requires three 3D FFTs (1 forward and 2 inverse) for each component of the considered vector quantity and they dominate theoretical time complexity of the method, which is of order $N^3 \log N^3$ (per time step).

In the context of this study, pseudo-spectral method proves to be particularly effective method.
Due to its use of spectral space and Fourier transforms it allows to easily incorporate large-scale forcing required to maintain homogeneous isotropic turbulence.
For the same reason, it is also well-suited to be used with LES method, especially in a form that is proposed in this study (see Section \ref{sc:ch1.les}), since velocity field filtering is defined in terms of convolution operation that becomes trivial to apply in Fourier space.
Moreover, Fourier transforms that are repeatedly applied to fluid velocity field are consistent with periodic boundary conditions.
Also, even though the velocity field is mainly integrated in the spectral space, it is also transferred back to the physical space (as opposed to purely spectral methods) and can be further used to calculate interactions between fluid flow and particles localised by their physical coordinates.
Finally, pseudo-spectral method is known for its stability (if Courant-Friedrichs-Lewy number does not exceed 0.3) and superior computational accuracy for homogeneous isotropic turbulence \parencite{Peng2009}.
On the other hand, frequent use of Fourier transforms makes it difficult to be effectively implemented for massively parallel execution, as such operations require global access to data across computational subdomains.
Such overheads for data transfer between processes rapidly increase with grid size, therefore it is important to find methods of reducing grid sizes as pseudo-spectral method does not seem to scale well performance-wise beyond grids of size $N=1024$.

@@

\begin{figure}[h]
\centering
\includegraphics[width=10cm]{figures/1-01_forcing.pdf}
\caption{
The basic idea of deterministic forcing scheme: energy is supplied on large scales by fixing it for small wavenumbers (bold line), then it is transferred following Kolmogorov energy cascade ($E(k) \propto k^{-5/3}$) in inertial subrange, to finally reach smallest dissipative eddies represented by largest resolved wavenumbers (when $k_{\max} \eta \sim 1 = 10^0$, diverging dashed lines) that have largest influence on measured particle statistics at contact distances.
Both axes are using standard normalisations (see Section \ref{ssc:ch2.flow.spec}) and logarithmic scale, that makes $k^{-5/3}$ curves appear as straight lines.}
\label{fig:forcing}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=17cm]{figures/1-02_dns-les-grids.pdf}
\caption{
Conceptual juxtaposition of computational grids for DNS and LES.
In LES, the subgrid-scale modelling parameterises small-scale turbulence that is directly resolved by $4 \times 4 \times 4$ block of cells in DNS.
}
\label{fig:dns-les-grids}
\end{figure}

In this study, that may be considered the continuation of \textcite{Rosa2017}, a similar setup for LES method is employed \parencite[based on earlier work by][]{Jin2010}.
In particular, the closure of the subgrid term is based on the concept of the spectral-eddy viscosity, $\nu_{e}$.
In spectral space, the filtered velocity field $\mathbf{\hat{\bar{U}}}(\mathbf{k}, t)$ is governed by the following momentum equation \parencite{Pope2000}
\begin{equation}
\left\lbrace \frac{\partial}{\partial t} + [\nu + \nu_{e}(k|k_{c})]k^{2} \right\rbrace \mathbf{\hat{\bar{U}}}(\mathbf{k}, t) = \mathbf{P}(\mathbf{k}) \mathcal{F}(\mathbf{\bar{U}} \times \boldsymbol{\bar{\omega}}) + \hat{\mathbf{f}}(\mathbf{k}, t) ,
\label{eqn:les}
\end{equation}
where $\mathbf{P} \equiv \left\{ \mathbf{P}_{ij} \right\}$ is the projection tensor (see Equation \ref{eqn:proj-tensor}) and $\mathcal{F}$ -- Fourier transform.
By $k_{c}$ we denote cutoff wavenumber for filtering and its value is set using the same formula as maximal resolved wavenumber, $k_{\max}$, in DNS, i.e. $k_c = \left\lfloor (N - 3) / 2 \right\rfloor$ (for $N=64$, we have $k_c = 30$).
In this method of modelling turbulence a sharp spectral filter is used to separate filtered (resolved) and residual (modelled) velocity fields which are strictly separated from each other in the spectral space with $| \mathbf{k} | = k_c$ as the boundary.
Hence, modelled modes for $| \mathbf{k} | > k_c$ have no bearing on turbulence model for larger wavenumbers and all necessary information on filtered velocity field $\mathbf{\hat{\bar{U}}}(\mathbf{k}, t)$ is contained in wavenumbers smaller than $k_c$.
Conversely, filtered velocity field contains no information on residual (small-scale) motions, thus they have to be entirely modelled \parencite[p. 615]{Pope2000}.

>>>>

\usepackage[table, svgnames, dvipsnames]{xcolor}
\usepackage[framemethod=tikz]{mdframed}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{algorithm}
\usepackage[backend=biber,style=authoryear-comp,sorting=nyt,maxcitenames=2,uniquename=false,uniquelist=false,maxbibnames=5,maxsortnames=1]{biblatex}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage[hidelinks,allcolors=black]{hyperref}
\usepackage{upgreek}

% requires LuaLaTeX
% \usepackage[nosingleletter]{impnattypo}

\renewcommand{\arraystretch}{1.1}

@@

\keywords{obliczeniowa mechanika płynów, przepływ turbulentny, przepływ wielofazowy, mikrofizyka chmur, krople, obliczenia dużej mocy, równania Naviera-Stokesa, homogeniczna izotropowa turbulencja, statystyki zderzeniowe kropel, metoda pseudo-spektralna, bezpośrednia symulacja numeryczna (DNS), metoda dużych wirów (LES), dwustronne sprzeżenie pędu, efekty grawitacyjne}

@@

\streszczenie{
Przepływy turbulentne z cząstkami fazy rozproszonej zachodzą nieustannie w środowisku naturalnym oraz znajdują liczne zastosowania w procesach przemysłowych.
Ważnym tego przykładem są chmury atmosferyczne, które składają się z ogromnej ilości małych kropel wody i kryształków lodu.
Wyjaśnienie wzajemnego oddziaływania powietrza (płynu) i aerozolu (cząstek) jest istotne w kontekście poznania mechanizmów rządzących powstawaniem i przebiegiem zjawisk atmosferycznych.
Wiedza w tym zakresie jest użyteczna w obszarze meteorologii i służy do opracowywania coraz to bardziej realistycznych modeli powstawania opadu z kropel chmurowych.
To z kolei prowadzi do podniesienia dokładności i sprawdzalności numerycznych prognoz pogody.
Jednym ze sposobów badania przepływów dyspersyjnych są symulacje numeryczne. 
\newline \indent
Powszechnie stosowanym do tego celu narzędziem, zapewniającym wysoką dokładność, są tzw. bezpośrednie symulacje numeryczne (direct numerical simulation, DNS). 
Za pomocą symulacji DNS można pozyskiwać statystyki zderzeniowe cząstek, które to są użyteczne jako parametry w modelach obejmujących większe skale przestrzenne.
Symulacje DNS są jednak kosztowne obliczeniowo, co ogranicza ich stosowalność do siatek o średnich rozmiarach (${\sim 2048^3}$ węzłów -- przy współcześnie dostępnych zasobach obliczeniowych).
DNS nie pozwalają modelować turbulencji charakteryzowanej wyższymi liczbami Reynoldsa (${R_{\lambda} \sim 10^4}$), czyli takiej która obserwowana jest w powietrzu atmosferycznym. \newline \indent
Metoda dużych wirów (large-eddy simulation, LES) stanowi alternatywę dla DNS.
W tej metodzie drobnoskalowa turbulencja nie jest bezpośrednio rozwiązywana, ale jest parametryzowana przy użyciu modelu podskalowego.
Umożliwia to modelowanie przepływów turbulentnych o większej liczbie Reynoldsa przy użyciu siatek o mniejszej liczbie węzłów.
Ograniczenie złożoności obliczeniowej odbywa się jednak kosztem mniejszej precyzji i fizycznej dokładności.
Niniejsza praca stanowi porównanie obu metod (DNS i LES), zarówno w kwestii fizycznej dokładności, jak i wydajności, skupiając się w szczególności na symulacjach z dwustronnym sprzężeniem pędu pomiędzy płynem i cząstkami. 
Zostało wykazane, że LES stanowi obiecującą odpowiedź na ograniczenia DNSu, pomimo pewnych niedokładności wynikających z~filtrowania turbulencji w najmniejszych skalach.
Różnice te, w większości przypadków, stają się mniejsze, gdy uwzględni się dwustronne sprzężenie pędu, w szczególności dla cząstek poddanych działaniu grawitacji.
\newline \indent
Przeprowadzone zostały również analizy kosztów obliczeniowych obu metod.
Choć zastosowanie LES zapewnia ogromną oszczędność w obliczeniach związanych z turbulentnym przepływem płynu, to jednak koszt ten znacząco rośnie wraz z liczbą cząstek.
Niezbędna ilość symulowanych cząstek może zaś być duża, gdyż efekty dwustronnego sprzężenia pędu objawiają się wyraźniej w układach z względnie wysokim stosunkiem masy cząstek i płynu (pomiędzy $0.1$~i~$1$).
Obliczenia odnoszące się do dynamiki cząstek stanowią więc bardziej istotne ograniczenie dla LESu niż dla DNSu, gdzie złożoność obliczeń związanych z~płynem jest czynnikiem dominującym.
Ponadto zidentyfikowano i przeanalizowano szereg parametrów, które pośrednio wpływają na czas trwania symulacji, takie jak np. promienie kropel oraz~działanie grawitacji.
Na koniec przedstawiono także wpływ preferencyjnego koncentrowania się cząstek na wariancję ich rozkładu pomiędzy równolegle wykonywanymi procesami.
Ten czynnik ma również istotne znaczenie w kontekście całkowitej wydajności symulacji.
}

@@

Many processes that involve fluid flows, whether laminar or turbulent, are composed of~more than one phase.
Here, by phase, we understand regions of the system with the same properties that can be clearly separated by some interface from others.
We refer to such flows as multi-phase, or two-phase, in a common case where the number of considered phases is limited to two.
These phases often are of different states of matter, e.g. gas--liquid for air bubbles flowing in the water pipe, but systems with two immiscible liquids, such as water and crude oil in pipelines, are also studied \parencite{Brauner2003}.
In many cases, one of the phases may be treated as continuous (referred to as the continuous phase, or the carrier fluid), while the other remains in separated chunks.
We refer to such chunks as the disperse phase, and examples may include air bubbles, water droplets, solid particulate matter suspended in~a~ fluid, etc.
Often enough, elements of disperse phase may be approximated as point particles, i.e. material points that are conceptually representing small spheres of given radii, that are considered insoluble and immiscible (that is, they cannot exchange mass with the carrier fluid, suddenly appear, or be destroyed).
Such approximation is especially accurate when disperse phase consists of small droplets or bubbles of one fluid surrounded by another, as the surface tension naturally preserves spherical shape of disperse elements.
Cloud droplets, in particular, remain spherical since the capillary pressure is several orders of magnitude larger than the local fluid shear stress induced by the disturbance flow associated with their motion.
Furthermore, \textcite[Figure 1, therein]{Balachandar2010} point out that for such approximation to be valid, particle response times have to be greater than Kolmogorov time scale of background turbulent flow. 
We shall refer to two-phase flows with disperse phase approximated by point particles as particle-laden.

@@

The modelling of atmospheric clouds is a broad domain that has to deal with wide range of length scales, where different structures and physical processes are important and have to be directly represented, while others need to be parameterised due to constraints in computing power.
Such tradeoff between computational complexity and physical fidelity of simulations gives rise to the plethora of methods used depending on the respective scale.
Simplified picture of that situation is presented in Figure \ref{fig:atmo-scales}.
At the largest scales we have global climate models (GCM) and global weather models (GWM, e.g. Integrated Forecasting System, Global Forecasting Systems, ICON), where computation domain  often spans the entire Earth (${\sim 10^{7}}$~m) and each cell has (horizontal) dimensions ranging from $10$ to $100$ kilometres.
We~also consider mesoscale models that are mainly used for weather prediction in smaller domains that usually cover regions, countries, or continents ($\sim 10^{5} \text{--} 10^{6}$ m) and grid spacing is in the order of kilometres.
For instance, maximal resolution of mesoscale model COSMO used for weather forecasting in Poland is $2.8$ km \parencite{Doms2021}.
In this context it is worth adding that a small cloud with a volume of $1 \; \text{km}^3$ may contain $10^{17}$ or more droplets.
Both types of models, due to relatively large grid cell sizes, cannot fully resolve even fairly large cloud structures and depend on a lot of parameterisations.
In the middle ground, we have cloud resolving models (CRM; domain sizes---kilometres, cell sizes---metres) that cannot fully resolve all turbulent scales of air or exactly model behaviour of every droplet, thus necessary parameterisations are employed, such as large-eddy simulations (LES) and super-droplet method \parencite{Arabas2013}.
CRM is commonly employed for research studies focused on the effects of microphysical processes affecting entire clouds, their formation and evolution, precipitation, etc.
For that reason, CRMs are used to refine cloud parameterisations used in GCM and mesoscale models.

\begin{figure}[ht]
\centering
\includegraphics[width=13cm]{figures/0-01_atmo-scales.png}
\caption{
Hierarchy of atmospheric models and the scales of atmospheric motion reproduced from \textcite{Morrison2020} with author's permission.
Coloured rectangles represent types of simulations commonly used when working with respective length scales, and comments emphasise transition from directly resolving most elements of the model (left) to including more and more parameterisations with the increase of scale (right).
Magenta rectangle positioned on the axis was added to emphasise range of scales that corresponds to simulations discussed in this thesis.
}
\label{fig:atmo-scales}
\end{figure}

@@

Recent studies, however, have shown that direct numerical simulations are close to their limits due to prohibitively large computational costs.
\textcite{Rosa2013,Rosa2016} used grids of up to $1024^3$ nodes, which seems to be the reasonable limit for most studies, although there were studies using even larger grids---e.g. $2048^3$ nodes in \textcite{Ireland2016} and \textcite{Matsuda2021}; for flows without particles grids of size $8192^3$ were used \parencite{Buaria2019} and more recently even up to $12\,288^{3}$ \parencite{Buaria2020}.
It is important to note that grid size is crucial for turbulent flow simulations since it affects the range of spatial scales that can be resolved---from the smallest eddies with sizes close to the spatial grid spacing, to the largest ones of sizes typically about $1/3$ of the computational domain (this limitation is due to a~numerical requirement that the correlation of fluid velocity at distances comparable to the domain size must vanish).
Such range between largest and smallest resolvable scales can be expressed using Reynolds numbe, which is commonly interpreted as a measure of magnitude of turbuence.
Examples of values of the Taylor microscale (that is, considering only eddies in inertial subrange) Reynolds number, $R_{\lambda}$, attained in previous high-resolution studies are $R_{\lambda} = 499$ for $1024^3$ grid \parencite{Rosa2013,Rosa2016}; $R_{\lambda} = 597$ \parencite{Ireland2016}, and $R_{\lambda} = 531$ \parencite{Matsuda2021} for $2048^3$ grids; $R_{\lambda} = 650$ for $8192^3$ \parencite{Buaria2019}; and $R_{\lambda} = 1300$ for $12\,288^3$ grid.
It is desirable, however, to attain higher values of $R_{\lambda}$ to simulate conditions that are present in typical atmospheric clouds, where $R_{\lambda}$ is known to attain values between $10^{3}$ and $10^{4}$.
What is more, the increase of the Taylor microscale Reynolds number with the number of grid nodes in one dimension, $N$, is sublinear (comparison of simulation data from various studies shows that $R_{\lambda} \propto N^{2/3}$, see \cite{Wang2009}, Figure 1 therein, which agrees with theoretical estimates for pseudo-spectral method in \cite{Pope2000}, Equation 9.8 therein).
Thus, when we double $N$ we only increase $R_{\lambda}$ by at most $50\% \text{--} 60\%$, while total number of grid nodes, $N^3$, increases by the factor of $8$.
The strategy of increasing grid sizes of DNS to model somewhat wider range of turbulent scales currently faces not only resource limitations, but is also hindered by the diminishing returns in the long run.
This situation calls for an alternative solution.

@@

Firstly, the accuracy, or physical fidelity, of LES is analysed.
Establishing exact measures of accuracy for collisional statistics is a challenging task because of experimental constraints, as simultaneous measurement of positions and velocities of large amounts of small droplets faces technical limitations, although considerable advances have been made recently in that field \parencite{Yavuz2018}.
Moreover, homogeneous and isotropic model of turbulence is an idealisation that could not be exactly replicated in actual physical conditions.
For these reasons it is difficult to obtain experimental results for validation, and due to the complexity of the problem at hand, no analytical solutions are available as well.
On the other hand, difficulties in obtaining experimental data makes estimations of these statistics \emph{in silico} even more valuable.
Thus, as in previous studies, the accuracy comparison will mostly focus on qualitative and quantitative differences between results obtained by DNS and LES.
The results of DNS are used as a baseline, under the assumption that this method is thoroughly studied and frequently used, and thus provides accurate estimates of parameters in question.
Hence, results obtained using LES are compared against them, in order to consider feasibility of LES as a~workable replacement for DNS.
The further analysis will focus mainly on simulations under two-way momentum coupling.

@@

The numerical integration of the Navier-Stokes equations is performed using the pseudo-spectral method which is well-established for such simulations.
It was first introduced by \textcite{Orszag1972} and later developed with parallel execution in mind and for application to particle-laden flows (e.g. see \textcite{Wang1993,Ayala2014,Parishani2015}).
It involves transformations of the velocity field between its representations in physical space, $\textbf{U}(\textbf{x}, t)$, and spectral (Fourier) space, $\hat{\textbf{U}}(\textbf{k}, t)$, in order to perform some operations in the space that is most favourable.
Detailed description of that method may be found in Appendix \ref{app:psm}.
Note, that performing Fourier transforms (implemented as FFT) is the most computationally demanding part of the method.
In the version of pseudo-spectral method used here, every iteration requires three 3D FFTs (1 forward and 2 inverse) for each component of the considered vector quantity and they dominate theoretical time complexity of the method, which is of order $N^3 \log N^3$ (per time step).

In the context of this study, pseudo-spectral method proves to be particularly effective method.
Due to its use of spectral space and Fourier transforms it allows to easily incorporate large-scale forcing required to maintain homogeneous isotropic turbulence.
For the same reason, it is also well-suited to be used with LES method, especially in a form that is proposed in this study (see Section \ref{sc:ch1.les}), since velocity field filtering is defined in terms of convolution operation that becomes trivial to apply in Fourier space.
Moreover, Fourier transforms that are repeatedly applied to fluid velocity field are consistent with periodic boundary conditions.
Also, even though the velocity field is mainly integrated in the spectral space, it is also transferred back to the physical space (as opposed to purely spectral methods) and can be further used to calculate interactions between fluid flow and particles localised by their physical coordinates.
Finally, pseudo-spectral method is known for its stability (if~Courant-Friedrichs-Lewy number does not exceed 0.3) and superior computational accuracy for homogeneous isotropic turbulence \parencite{Peng2009}.
On the other hand, frequent use of Fourier transforms makes it difficult to be effectively implemented for massively parallel execution, as such operations require global access to data across computational subdomains.
Such overheads for data transfer between processes rapidly increase with grid size, therefore it is important to find methods of reducing grid sizes as pseudo-spectral method does not seem to scale well performance-wise beyond grids of size $N=1024$.

@@

\begin{figure}[ht]
\centering
\includegraphics[width=10cm]{figures/1-01_forcing.pdf}
\caption{
The basic idea of deterministic forcing scheme: energy is supplied on large scales by fixing it for small wavenumbers (bold line), then it is transferred following Kolmogorov energy cascade ($E(k) \propto k^{-5/3}$) in inertial subrange, to finally reach smallest dissipative eddies represented by largest resolved wavenumbers (when $k_{\max} \eta \sim 1 = 10^0$, diverging dashed lines) that have largest influence on measured particle statistics at contact distances.
Both axes are using standard normalisations (see Section \ref{ssc:ch2.flow.spec}) and logarithmic scale, that makes $k^{-5/3}$ curves appear as straight lines.}
\label{fig:forcing}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=17cm]{figures/1-02_dns-les-grids.pdf}
\caption{
Conceptual juxtaposition of computational grids for DNS and LES.
In LES, the subgrid-scale modelling parameterises small-scale turbulence that is directly resolved by $4 \times 4 \times 4$ block of cells in DNS.
}
\label{fig:dns-les-grids}
\end{figure}

In this study, that may be considered the continuation of \textcite{Rosa2017}, a~similar setup for LES method is employed \parencite[based on earlier work by][]{Jin2010}.
In particular, the closure of the subgrid term is based on the concept of the spectral-eddy viscosity, $\nu_{e}$.
In spectral space, the filtered velocity field $\mathbf{\hat{\bar{U}}}(\mathbf{k}, t)$ is governed by the following momentum equation \parencite{Pope2000}
\begin{equation}
\left\lbrace \frac{\partial}{\partial t} + [\nu + \nu_{e}(k|k_{c})]k^{2} \right\rbrace \mathbf{\hat{\bar{U}}}(\mathbf{k}, t) = \mathbf{P}(\mathbf{k}) \mathcal{F}(\mathbf{\bar{U}} \times \boldsymbol{\bar{\omega}}) + \hat{\mathbf{f}}(\mathbf{k}, t) ,
\label{eqn:les}
\end{equation}
where $\mathbf{P} \equiv \left\{ \mathbf{P}_{ij} \right\}$ is the projection tensor (see Equation \ref{eqn:proj-tensor}) and $\mathcal{F}$ -- Fourier transform.
By $k_{c}$ we denote cutoff wavenumber for filtering and its value is set using the same formula as maximal resolved wavenumber, $k_{\max}$, in DNS, i.e. $k_c = \left\lfloor (N - 3) / 2 \right\rfloor$ (for $N=64$, we have $k_c = 30$).
In this method of modelling turbulence a sharp spectral filter is used to separate filtered (resolved) and residual (modelled) velocity fields which are strictly separated from each other in the spectral space with $| \mathbf{k} | = k_c$ as the boundary.
Hence, modelled modes for $| \mathbf{k} | > k_c$ have no bearing on turbulence model for larger wavenumbers and all necessary information on filtered velocity field $\mathbf{\hat{\bar{U}}}(\mathbf{k}, t)$ is contained in wavenumbers smaller than $k_c$.
Conversely, filtered velocity field contains no information on residual (small-scale) motions, thus they have to be entirely modelled \parencite[p. 615]{Pope2000}.
